\normalfont\spacedallcaps{Hazardous asteroids forecast via Markov random fields}

### 2021-12-03 Manuscript/manuscript.tex
PUB

\section*{Abstract} The machine learning algorithms provide a promising approach to classify and predict natural or social phenomena. Differently to the theoretical approach, in which the laws that describe particular phenomena are derived from general assumptions using a mathematical language (e.g., a symmetry, the principle of energy conservation, the three Newton laws, the second principle of thermodynamics \cite{feynmanlectures}), the machine learning algorithms start from very general assumptions/expressions and then their parameters are optimized with the likehood maximization\cite{russell2010artificial,murphy2012machine}. The first approach allows obtaining interpretable predictions, the second one good forecasts with a reduced effort. However, the toll that has to be paid in this case is the low interpretability \cite{russell2010artificial,murphy2012machine}. Since they provide the list of conditional independences/dependencies, the graphical methods represent a good compromise between the need for interpretability and the forecasts obtained without developing a general theory. In order to prove the validity of this statement, I considered a dataset for which the theory that interconnects its features was known: in particular, I chose the asteroids hazardousness as provided by CNEOS \cite{cneos+nasa} and published on Kaggle \cite{kaggle_dataset}. The outcomes of the probabilistic methods for predicting the asteroid hazardousness were compared with the ones provided by the theory and with the ones of Random Forest (RF), Support Vector Machines (SVM), and Logistic Regression and Quadric Discriminant analysis (QDA). The results show that the forecast performances of probabilistic methods are better than QDA and almost equal to the logistic regression but lower to RF and SVM. However, since the list of conditional dependencies correctly reflects the laws of celestial mechanics \cite{murray1999solar}, it can be said that in this case, the probabilistic methods provide an interpretable and correct explanation of their mechanism. Therefore, contrary to other machine learning algorithms, such methods can be fully validated scientifically.

@@

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as measurable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the ab-initio case, one can fully explain the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The drawback of this approach is that the underlying mechanism,  by which a forecast is preferred to another one, may not be interpretable. Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide conditional dependencies/independences, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the theory of the involved process is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contains the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. This paper is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset. Finally, the code used to produce the results here reported if provided in Appendix B.

@@

In this case, a first-order Markov chain is obtained,  where the transition tensor is of second-order \cite{wu2017markov}.  Given this formalism, we are interested in finding an intelligent way to plot such joint distribution intuitively: the graph theory answers this quest.  In particular, the absence of an edge that interconnects the random variables can be represented by nodes and the presence of conditional independence for two random variables.  Bayesian networks consider directed edges,  while Markov random fields (MRF) are only undirected.  Consequently,  while the concept of a topological ordering,  by which the parents n nodes are labelled lower to their children,  is well defined for the Bayesian network, MRF is not.  In order to solve this issue, it is helpful to consider the Hammersley-Clifford theorem as stated in \cite{murphy2012machine}:

@@

\section{Probabilistic models} The analysis with the graphical models was started by inspecting the relations between continuous variables in the dataset. Therefore, in this first step, the binary variable \textit{Hazardous} was excluded. For this purpose, among the different methods available, I considered the \textit{Graphical least absolute shrinkage and selection operator} GLASSO as implemented in the \textit{glasso} R package \cite{friedman2008sparse,glasso}. After different tests reported in Fig. \ref{GLASSO_convergence}, I considered as the final result the graph obtained with the value of $\rho$ (the one that penalizes further connections) equal to $0.3$. Such choice is motivated by the fact that the conditional dependences/independences stated by the Celestial Mechanics are correctly reproduced with this parameter. For instance, it can be seen that the diameter of the asteroids is independent of all features related to their motion. Such a result is meaningful since the volume/mass of the asteroids is largely lower than the mass of the earth.  Thus, as explained in Appendix A, there is no way the asteroid orbit can be modified by its mass. In addition, it can be seen that the Close approach Date and the Epoch date are linked to each other but in no way from the other features. This result is correct since the date and epoch are set with an arbitrary scale. Also, as expected, the Perihelion Epoch and Osculation Time are linked, but there is no dependence on the orbit parameters. Furthermore, the mean motion is conditionally independent with relative velocity, but correctly, this is dependent on perihelion distance.

@@

Let us now move on to the mixed interaction model, where the hazardousness feature was also considered.  The first model was calculated with the mgm package  \cite{mgm,haslbeck2015mgm} where the nodewise regression is used \cite{meinshausen2006high}. The $k$ parameter was set equal to two, and a cross-validation (CV) with ten folds was considered. The result is reported in Fig. \ref{mgm}. From this figure, it can be seen that the features that are conditionally dependant on the Hazardous features are: the absolute magnitude, the min orbit intersection, and the eccentricity.  The theory supports all these connections. The min orbit intersection is the main parameter for the hazardous value. On the other side, the eccentricity can be tough as a parameter that describes how near the celestial body moves at the perihelion (indeed, this parameter is correctly linked with the perihelion distance). The Absolute Magnitude is also a meaningful parameter for the hazard evaluation since, if the asteroid is too small, it will be destroyed by the earth atmosphere.  Furthermore, the parameters related to the diameters are also not connected with the orbital parameters. In addition, the mean velocity is not directly related to the mean motion, but there are the orbital parameters within as expected. Thus it can be said that the model produced is almost consistent with the astronomic laws. On this basis, the performances of the model in terms of confusion matrix, ROC (Receiver operating characteristic) curve and $\phi$ coefficient (also known as Matthews correlation coefficient ) were evaluated (for this purpose the Caret R package \cite{kuhn2008building,caret} was used). The confusion matrix obtained is reported in Fig. \ref{mgm_confusion}: the ROC curve is given in Fig. \ref{ROC_mgm} and it corresponds to a $\phi=0.6$. These performances will be commented on in the next section when the performances of other ML algorithms for the same dataset will be reported. Beside the mgm algorithm/package other approaches were also considered for the evaluation of the mixed interaction graphical model: the function \textit{mmod()} of gRim package \cite{hojsgaard2012graphical} and the \textit{minforest()} of gRapHD (which uses the minForest method) \cite{de2009high}. For both cases, a stepwise algorithm was used.  Their result are reported in Fig. \ref{mmod} and \ref{minforest}. It can be seen that in these models, the minimum diameter is linked with the hazardous feature, which is correct, but what is without a physical meaning is that this quantity is also linked to the features connected with the orbital parameters. A weak link with the orbital parameters is meaningful for the magnitude (as the mgm model for the mean anomaly).  The magnitude depends on the distance between the asteroids and the observer.   However, a direct link from the volume parameters (and thus from the mass) with the orbital ones is not acceptable. In principle, one can set these links as forbidden in the algorithm (blacklist), but in the author view, a model where the result is obtained without boundaries should be preferred to those obtained with a large number of boundaries. Therefore we reject the mmod and minforest ones favouring the mgm, which was obtained without boundaries.

@@

\section{Machine learning algorithms} The performances of mgm model will be now compared with the one obtained from other ML algorithms. These include the random-forest (RF, as implemented in the RandomForest package \cite{rfor}), the support vector machines (SVM, as implemented in the e1071 package \cite{dimitriadou2008misc}), the quadratic discriminant anchiedersialysis (QDA, as implemented in the MASS package \cite{MASin reverseS} and the logistic regression (as implemented in stats package \cite{stats}). Their performances are reported in Fig. \ref{CF_ML} and \ref{ROC_mgm} as well in the Tab. \ref{phi_values}. This comparison shows that the RF and the SVM outperform the mgm graphical method while the logistic one has similar performances. Thus, one can wonder about the advantage of using a graphical method instead of a random forest or an SVM since its performances seem lower. The answer is the interpretability of the model provided: the RF, at least, can provide variable rank importance as reported in Fig. \ref{RF_Importance} (note that apart from the Min Orbit intersection, there is a slight reshuffling in the feature importance to the one estain reverseblished by the MI in fig \ref{Mutual_information}. However, apart from this intepretation, the RF is a black-box as the SVM, the logistic regression and the QDA. Conversely, the probabilistic graphs, providing conditional independences, give the user an interpretable model whose properties can also be compared, discussed, and validated with the theory. Thus the model developed in this way allows a more scientific evaluation with respect to the black-box ones. In the author's view, this characteristic compensates for the lack of predictive power regarding the RF or the SVM methods.

@@

The GLASSO (with a penalizing parameter $\rho=0.3$) and the mgm algorithms provided graphical models that correctly reproduce the physical connections of the dataset feature. In particular, for both models, it can be seen that no one of the orbital parameters is conditionally dependent on the diameter of the asteroid, as stated by theory. Furthermore, the mgm model can provide forecasts about the hazardousness of the asteroids with almost equal or better performances than the logistic and QDA algorithms.  On the other hand, despite the mgm performances being lower than the RF and SVM, its interpretability is higher.  Such a feature compensates the mgm forecast performances gap. As a future outlook I consider to extend this analysis for physical phenomena for which there is no deterministic theory but only probabilistic one\footnote{Following the I.Prigogine \cite{prigogine2017non,prigogine1997end,nicolis1989exploring,prigogine1978time} argument it must be said that a pure deterministic theory, if exist, cannot be fully applied. Therefore, in principle, all possible theories should be considered probabilistic. This is because the initial conditions can not be known with a infinite precision. Therefore, if one also perfectly knows the equation of motion and can also solve them exactly (and in most of cases it is not possible) a source of error still persist and it propagates with the Lyapunov exponent. However as it was shown by Prigogine it a system is at equilibrium this instability can neglected. In this case a deterministic approach is meaningful }. For instance one can consider Earthquakes and Vulcanos Eruptions \cite{sornette1989self,sparks2003forecasting,sornette2006critical}

>>>>

\section*{Abstract} The machine learning algorithms provide a promising approach to classify and predict natural or social phenomena. Differently to the theoretical approach, in which the laws that describe particular phenomena are derived from general assumptions using a mathematical language (e.g., a symmetry, the principle of energy conservation, the three Newton laws, the second principle of thermodynamics \cite{feynmanlectures}), the machine learning algorithms start from very general assumptions/expressions and then their parameters are optimized with the likehood maximization\cite{russell2010artificial,murphy2012machine}. The first approach allows obtaining interpretable predictions, the second one good forecasts with a reduced effort. However, the toll that has to be paid in this case is the low interpretability \cite{russell2010artificial,murphy2012machine}. Since they provide the list of connections within the features, the graphical methods represent a good compromise between the need for interpretability and the forecasts obtained without developing a general theory. In order to prove the validity of this statement, I considered a dataset for which the theory that interconnects its features was known: in particular, I chose the asteroids hazardousness as provided by CNEOS \cite{cneos+nasa} and published on Kaggle \cite{kaggle_dataset}. The outcomes of the probabilistic methods for predicting the asteroid hazardousness were compared with the ones provided by the theory and with the ones of Random Forest (RF), Support Vector Machines (SVM), and Logistic Regression and Quadric Discriminant analysis (QDA). The results show that the forecast performances of probabilistic methods are better than QDA and almost equal to the logistic regression but lower to RF and SVM. However, since the list of connections correctly reflects the laws of celestial mechanics \cite{murray1999solar}, it can be said that in this case, the probabilistic methods provide an interpretable and correct explanation of their mechanism. Therefore, contrary to other machine learning algorithms, such methods can be fully validated scientifically.

@@

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as measurable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the ab-initio case, one can fully explain the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The drawback of this approach is that the underlying mechanism,  by which a forecast is preferred to another one, may not be interpretable. Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide the connection list, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the theory of the involved process is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contains the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. This paper is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset. Finally, the code used to produce the results here reported if provided in Appendix B.

@@

In this case, a first-order Markov chain is obtained,  where the transition tensor is of second-order \cite{wu2017markov}.  Given this formalism, we are interested in finding an intelligent way to plot such joint distribution intuitively: the graph theory answers this quest.  The nodes can be used to represent the random variables while the presence or the lack of edges can be used to represent the conditional indipendence \cite{murphy2012machine}.  Bayesian networks consider directed edges,  while Markov random fields (MRF) are only undirected.  Consequently,  while the concept of a topological ordering,  by which the parents n nodes are labelled lower to their children,  is well defined for the Bayesian network, MRF is not.  In order to solve this issue, it is helpful to consider the Hammersley-Clifford theorem as stated in \cite{murphy2012machine}:

@@

\section{Probabilistic models} The analysis with the graphical models was started by inspecting the relations between continuous variables in the dataset. Therefore, in this first step, the binary variable \textit{Hazardous} was excluded. For this purpose, among the different methods available, I considered the \textit{Graphical least absolute shrinkage and selection operator} GLASSO as implemented in the \textit{glasso} R package \cite{friedman2008sparse,glasso}. After different tests reported in Fig. \ref{GLASSO_convergence}, I considered as the final result the graph obtained with the value of $\rho$ (the one that penalizes further connections) equal to $0.3$. Such choice is motivated by the fact that the connections stated by the Celestial Mechanics are correctly reproduced with this parameter. For instance, it can be seen that the diameter of the asteroids is not connected to all features related to their motion. Such a result is meaningful since the volume/mass of the asteroids is largely lower than the mass of the earth.  Thus, as explained in Appendix A, there is no way the asteroid orbit can be modified by its mass. In addition, it can be seen that the Close approach Date and the Epoch date are linked to each other but in no way from the other features. This result is correct since the date and epoch are set with an arbitrary scale. Also, as expected, the Perihelion Epoch and Osculation Time are linked, but there is no dependence on the orbit parameters. Furthermore, the mean motion is not connected directly to relative velocity, but correctly, both are connected to the perihelion distance.

@@

Let us now move on to the mixed interaction model, where the hazardousness feature was also considered.  The first model was calculated with the mgm package  \cite{mgm,haslbeck2015mgm} where the nodewise regression is used \cite{meinshausen2006high}. The $k$ parameter was set equal to two, and a cross-validation (CV) with ten folds was considered. The result is reported in Fig. \ref{mgm}. From this figure, it can be seen that the features that are directly connected to the Hazardous feature are: the absolute magnitude, the min orbit intersection, and the eccentricity.  The theory supports all these connections. The min orbit intersection is the main parameter for the hazardous value. On the other side, the eccentricity can be tough as a parameter that describes how near the celestial body moves at the perihelion (indeed, this parameter is correctly linked with the perihelion distance). The Absolute Magnitude is also a meaningful parameter for the hazard evaluation since, if the asteroid is too small, it will be destroyed by the earth atmosphere.  Furthermore, the parameters related to the diameters are also not connected with the orbital parameters. In addition, the relative velocity is not directly related to the mean motion, but there are the orbital parameters within as expected. Thus it can be said that the model produced is almost consistent with the astronomic laws. On this basis, the performances of the model in terms of confusion matrix, ROC (Receiver operating characteristic) curve and $\phi$ coefficient (also known as Matthews correlation coefficient ) were evaluated (for this purpose the Caret R package \cite{kuhn2008building,caret} was used). The confusion matrix obtained is reported in Fig. \ref{mgm_confusion}: the ROC curve is given in Fig. \ref{ROC_mgm} and it corresponds to a $\phi=0.6$. These performances will be commented on in the next section when the performances of other ML algorithms for the same dataset will be reported. Beside the mgm algorithm/package other approaches were also considered for the evaluation of the mixed interaction graphical model: the function \textit{mmod()} of gRim package \cite{hojsgaard2012graphical} and the \textit{minforest()} of gRapHD (which uses the minForest method) \cite{de2009high}. For both cases, a stepwise algorithm was used.  Their result are reported in Fig. \ref{mmod} and \ref{minforest}. It can be seen that in these models, the minimum diameter is linked with the hazardous feature, which is correct, but what is without a physical meaning is that this quantity is also linked to the features connected with the orbital parameters. A weak link with the orbital parameters is meaningful for the magnitude (as the mgm model for the mean anomaly).  The magnitude depends on the distance between the asteroids and the observer.   However, a direct link from the volume parameters (and thus from the mass) with the orbital ones is not acceptable. In principle, one can set these links as forbidden in the algorithm (blacklist), but in the author view, a model where the result is obtained without boundaries should be preferred to those obtained with a large number of boundaries. Therefore we reject the mmod and minforest ones favouring the mgm, which was obtained without boundaries.

@@

\section{Machine learning algorithms} The performances of mgm model will be now compared with the one obtained from other ML algorithms. These include the random-forest (RF, as implemented in the RandomForest package \cite{rfor}), the support vector machines (SVM, as implemented in the e1071 package \cite{dimitriadou2008misc}), the quadratic discriminant anchiedersialysis (QDA, as implemented in the MASS package \cite{MASin reverseS} and the logistic regression (as implemented in stats package \cite{stats}). Their performances are reported in Fig. \ref{CF_ML} and \ref{ROC_mgm} as well in the Tab. \ref{phi_values}. This comparison shows that the RF and the SVM outperform the mgm graphical method while the logistic one has similar performances. Thus, one can wonder about the advantage of using a graphical method instead of a random forest or an SVM since its performances seem lower. The answer is the interpretability of the model provided: the RF, at least, can provide variable rank importance as reported in Fig. \ref{RF_Importance} (note that apart from the Min Orbit intersection, there is a slight reshuffling in the feature importance to the one estain reverseblished by the MI in fig \ref{Mutual_information}. However, apart from this intepretation, the RF is a black-box as the SVM, the logistic regression and the QDA. Conversely, the probabilistic graphs, providing the list of connections among the random variables, give the user an interpretable model whose properties can also be compared, discussed, and validated with the theory. Thus the model developed in this way allows a more scientific evaluation with respect to the black-box ones. In the author's view, this characteristic compensates for the lack of predictive power regarding the RF or the SVM methods.

@@

The GLASSO (with a penalizing parameter $\rho=0.3$) and the mgm algorithms provided graphical models that correctly reproduce the physical connections of the dataset feature. In particular, for both models, it can be seen that no one of the orbital parameters is conneted to the diameter of the asteroid, as stated by theory. Furthermore, the mgm model can provide forecasts about the hazardousness of the asteroids with almost equal or better performances than the logistic and QDA algorithms.  On the other hand, despite the mgm performances being lower than the RF and SVM, its interpretability is higher.  Such a feature compensates the mgm forecast performances gap. As a future outlook I consider to extend this analysis for physical phenomena for which there is no deterministic theory but only probabilistic one\footnote{Following the I.Prigogine \cite{prigogine2017non,prigogine1997end,nicolis1989exploring,prigogine1978time} argument it must be said that a pure deterministic theory, if exist, cannot be fully applied. Therefore, in principle, all possible theories should be considered probabilistic. This is because the initial conditions can not be known with a infinite precision. Therefore, if one also perfectly knows the equation of motion and can also solve them exactly (and in most of cases it is not possible) a source of error still persist and it propagates with the Lyapunov exponent. However as it was shown by Prigogine it a system is at equilibrium this instability can neglected. In this case a deterministic approach is meaningful }. For instance one can consider Earthquakes and Vulcanos Eruptions \cite{sornette1989self,sparks2003forecasting,sornette2006critical}
